---
title: "Chapter 2"
output: html_document
---

```{r load_tidyverse_edr}
library(tidyverse)
library(edr)
```

## 2.1. Programming Basics and Package Installation

### Listing 2.1. Assigning a single number to `a_number`.

```{r}
a_number <- 5

a_number
```

### Listing 2.2. Assigning a vector of three numbers to `three_numbers`.

```{r}
three_numbers <- c(1, 2, 3)

three_numbers
```

### Listing 2.3. Assigning a character string to `one_word`.

```{r}
one_word <- "hello"

one_word
```

### Listing 2.4. Replacing the `one_word` variable with a different value.

```{r}
one_word <- "hi"

one_word
```


## 2.2. Using **dplyr** to Transform Data

### Listing 2.5. Loading the **tidyverse** and **edr** packages.

```{r load_tidyverse_edr, eval=FALSE}
library(tidyverse)
library(edr)
```

### Listing 2.6. The `sw` dataset can be printed to the console by using `sw`.

```{r}
sw
```

### Listing 2.7. The `filter()` function, and filtering by a single value in the `species` column of the `sw` dataset.

```{r sw_filter_droid, paged.print=FALSE}
filter(sw, species == "Droid")
```

### Listing 2.8. Filtering to only keep rows where the `height` column has values greater than 220.

```{r sw_filter_height_gt_220, paged.print=FALSE}
filter(sw, height > 220)
```

### Listing 2.9. Filtering characters by either `height` above 210, or, `mass` above 120.

```{r sw_height_mass, paged.print=FALSE}
filter(sw, height > 210 | mass > 120)
```

### Listing 2.10. Get all `Droid` and `Human` characters (method 1 with a [expr | expr] construction). 

```{r sw_species_or, paged.print=FALSE}
filter(sw, species == "Droid" | species == "Human")
```

### Listing 2.11. Get all `Droid` and `Human` characters (method 2 with `[colname %>% c(...)]` construction).

```{r sw_species_in, paged.print=FALSE}
filter(sw, species %in% c("Droid", "Human"))
```

### Listing 2.12. Arranging characters by increasing height.

```{r sw_arrange_height, paged.print=FALSE}
arrange(sw, height)
```

### Listing 2.13. Arranging characters by hair color and then by gender.

```{r sw_arrange_twice, paged.print=FALSE}
arrange(sw, hair_color, gender)
```

### Listing 2.14. Arranging by descending `hair_color`, then by ascending `gender`.

```{r sw_arrange_twice_desc, paged.print=FALSE}
arrange(sw, desc(hair_color), gender)
```

### Listing 2.15. Selecting the `name`, `gender`, and `species` columns.

```{r sw_select, paged.print=FALSE}
select(sw, name, gender, species)
```

### Listing 2.16. Selecting the `gender`, `species`, and `names` columns (in that order).

```{r sw_select_revised, paged.print=FALSE}
select(sw, gender, species, name)
```

### Listing 2.17. Selecting the `name` column and any additional columns ending with "s".

```{r sw_select_ends_with, paged.print=FALSE}
select(sw, name, ends_with("s"))
```

### Listing 2.18. Selecting the `name` column and any additional columns containing a "_" character.

```{r sw_select_contains, paged.print=FALSE}
select(sw, name, contains("_"))
```

### Listing 2.19. Selecting three specific columns and then `everything()` else after that.

```{r sw_select_w_everything, paged.print=FALSE}
select(sw, name, homeworld, gender, everything())
```

### Listing 2.20. Create the `sw_small` table containing just three columns.

```{r make_sw_smaller, paged.print=FALSE}
sw_small <- select(sw, name, height, mass)
sw_small
```

### Listing 2.21. Add the `bmi` column to `sw_small` using the `mutate()` function.

```{r mutate_bmi, paged.print=FALSE}
mutate(sw_small, bmi = mass / (height / 100)^2)
```

### Listing 2.22. Add the `bmi` column as before and then create `bmi_rnd`.

```{r mutate_bmi_round, paged.print=FALSE}
mutate(
  sw_small,
  bmi = mass / (height / 100)^2,
  bmi_rnd = round(bmi, 0)
)
```

### Listing 2.23. Grouping by the `species` variable and summarizing to get the mean mass.

```{r group_by_summarize_mean, paged.print=FALSE}
# Create grouped data; where the table is grouped by `species`
by_species <- group_by(sw, species)

# Create a data summary that gets the mean mass by `species`
summarize(by_species, avg_mass = mean(mass))
```

### Listing 2.24. A slight modification to the summary expression to deal with missing (`NA`) values.

```{r summarize_mean_na_rm, paged.print=FALSE}
summarize(by_species, avg_mass = mean(mass, na.rm = TRUE))
```

### Listing 2.25. Grouping by two columns and obtaining a summary with two new columns.

```{r sw_summarize_two_column_groups, paged.print=FALSE}
# Create grouped data; grouping by `species` and `gender`
grouping <- group_by(sw, species, gender)

# Summarize to get average mass and average height
summarize(
  grouping,
  avg_mass = mean(mass, na.rm = TRUE), 
  avg_height = mean(height, na.rm = TRUE)
)
```

### Listing 2.26. Bringing it all together with multiple **dplyr** functions and assigning each step to a different object.

```{r dplyr_stmts, paged.print=FALSE}
# Get average BMI values for humans across the different worlds

humans <- filter(sw, species == "Human")

humans_bmi <- mutate(humans, bmi = mass / (height / 100)^2)

humans_bmi_by_homeworld <- group_by(humans_bmi, homeworld)

humans_avg_bmi_by_homeworld <- 
  summarize(humans_bmi_by_homeworld, avg_bmi = mean(bmi, na.rm = TRUE))

humans_avg_bmi_by_homeworld_sorted <-
  arrange(humans_avg_bmi_by_homeworld, desc(avg_bmi))

humans_avg_bmi_by_homeworld_sorted
```

### Listing 2.27. Bringing it all together with multiple **dplyr** functions: this time using the pipe operator.

```{r dplyr_piped, paged.print=FALSE}
bmi_sorted <-
  sw %>%
  filter(species == "Human") %>%
  mutate(bmi = mass / (height / 100)^2) %>%
  group_by(homeworld) %>%
  summarize(avg_bmi = mean(bmi, na.rm = TRUE)) %>%
  arrange(desc(avg_bmi))

bmi_sorted
```

### Listing 2.28. Using pseudocode to demonstrate the readability of the code in the previous listing.

```{r}
# Get a table of average BMI values for humans across
# the different worlds in a single, piped expression;
# The following is written as pseudocode
# (don’t run it, just read it)

# {assigning to bmi_sorted} <-
#   {start with the starwars table} %{then}%
#   {keep only the rows where species is “Human”} %{then}%
#   {create a new column called bmi using a calculation} %{then}%
#   {create a group for each unique value in homeworld} %{then}%
#   {for each group create a summary table (one col: avg_bmi)} %{then}%
#   {order the rows by decreasing avg_bmi value}
# 
# {view bmi_sorted}
```


## 2.3. Creating our own Tabular Data

### Listing 2.29. Using `tibble()` with equal-length vectors to make a tibble.

```{r tibble_all_n, paged.print=FALSE}
tibble( 
  a = c(3, 5, 2, 6),
  b = c("a", "b", "g", "b")
)
```

### Listing 2.30. Using `tibble()` with two vectors: one of length 1 and the other of length 4.

```{r tibble_n_1, paged.print=FALSE}
tibble(
  a = 3,
  b = c("a", "b", "g", "b")
)
```

### Listing 2.31. Using `tibble()` with two vectors that contain `NA` values.

```{r tibble_na_vals, paged.print=FALSE}
tibble(
  a = c(3, 5, 2, NA),
  b = c("a", NA, "g", "b")
)
```

### Listing 2.32. Using a single-length vector with an `NA` value in `tibble()`.

```{r tibble_na_col, paged.print=FALSE}
tibble(
  a = NA,
  b = c("a", "b", "g", "b")
)
```

### Listing 2.33. Using a single-length vector with an `NA` value in `tibble()`.

```{r tibble_na_character, paged.print=FALSE}
tibble(
  a = NA_character_,
  b = c("a", "b", "g", "b")
)
```

### Listing 2.34. Creating a tibble using the `tribble()` function.

```{r tribble_use, paged.print=FALSE}
tribble(
  ~a, ~b,
  3,  "a",
  5,  "b",
  2,  "g",
  6,  "b",
)
```
