---
title: "Chapter 9"
output: html_document
---

```{r load_packages}
library(tidyverse)
library(edr)
```

### Listing 9.1. Creating a three-element, named list object

```{r named_list_of_3_elements, paged.print=FALSE}
a_list <- 
  list(
    a = c(1:5, 8:3, 2),
    b = tibble(x = 1:3, y = rep(TRUE, 3)),
    c = list(A = 1:3, B = LETTERS[20:26])
  )

a_list
```

### Listing 9.2. Examining the structure of a list object with `str()`

```{r structure_of_a_list}
str(a_list)
```

### Listing 9.3. Getting the names of the list components with `names()`

```{r names_in_a_list}
names(a_list)
```

### Listing 9.4. Getting the first element of `a_list` with `[ ]`

```{r a_list_element_single_brackets}
a_list[1]
```

### Listing 9.5. Getting the first element of `a_list` with `[[ ]]`

```{r a_list_element_double_brackets}
a_list[[1]]
```

### Listing 9.6. Creating a three-element, unnamed list object

```{r unnamed_list_of_3_elements, paged.print=FALSE}
b_list <- 
  list(
    c(1:5, 8:3, 2),
    tibble(x = 1:3, y = rep(TRUE, 3)),
    list(A = 1:3, B = LETTERS[20:26])
  )

b_list
```

### Listing 9.7. Examining the structure of the `b_list` object with `str()`

```{r structure_of_the_b_list}
str(b_list)
```

### Listing 9.8. Giving the unnamed list (`b_list`) some names

```{r b_list_gets_names, paged.print=FALSE}
names(b_list) <- c("a", "b", "c")

b_list
```

### Listing 9.9. Modifying the tibble component of `b_list` requires subsetting with `b_list$b`

```{r modifying_the_b_list_tibble, paged.print=FALSE}
b_list$b <-
  b_list$b %>%
  mutate(z = c("a", "b", "c"))

b_list
```

### Listing 9.10. Combining three named lists into a larger list with `c()` (reproducing the `a_list` object)

```{r combining_lists_with_c, paged.print=FALSE}
a_1 <- list(a = c(1:5, 8:3, 2))
a_2 <- list(b = tibble(x = 1:3, y = rep(TRUE, 3)))
a_3 <- list(c = list(A = 1:3, B = LETTERS[20:26]))

a_list <- c(a_1, a_2, a_3)

a_list
```

### Listing 9.11. Removing the second element of the `a_list` object using `NULL`.

```{r removing_element_with_NULL, paged.print=FALSE}
a_list$b <- NULL

a_list
```

### Listing 9.12. Removing the first element of the `a_list` object using `NULL`, this time with an index value.

```{r removing_element_with_NULL_and_index, paged.print=FALSE}
a_list[1] <- NULL

a_list
```

### Listing 9.13. A named list that is suitable for transformation into a data frame or tibble.

```{r, list_to_transform_to_table, paged.print=FALSE}
list_df <- list(col_1 = c(4, 5, 6), col_2 = c("x", "y", "z"))

list_df
```

### Listing 9.14. Transforming the `list_df` list into a data frame with `as.data.frame()`.

```{r, transform_as_data_frame, paged.print=FALSE}
df_from_list <- as.data.frame(list_df, stringsAsFactors = FALSE) 

df_from_list
```

### Listing 9.15. Transforming the `list_df` list into a tibble with `as_tibble()`.

```{r, transform_as_tibble, paged.print=FALSE}
tbl_from_list <- as_tibble(list_df) 

tbl_from_list
```

### Listing 9.16. Transforming the `tbl_from_list` tbl back to a list with `as.list()`.

```{r, transform_tibble_to_list, paged.print=FALSE}
list_from_tbl <- as.list(tbl_from_list) 

list_from_tbl
```

### Listing 9.17. Example of a function (`get_descriptive_stats()`) that returns a list object.

```{r, get_descriptive_stats}
get_descriptive_stats <- function(x, na.rm = TRUE) {
  
  list(
    values = x,
    mean = mean(x, na.rm = na.rm),
    sd   = sd(x, na.rm = na.rm),
    min  = min(x, na.rm = na.rm),
    max  = max(x, na.rm = na.rm),
    rank = rank(x)
  )
}

stats <- get_descriptive_stats(c(2.3, 8.1, 5.5, 3.9, 6.0, 2.1, 8.5))

stats
```

### Listing 9.18. Example of a function (`add_percentiles()`) that takes a list object as input and returns an augmented version of the input list.

```{r, add_percentiles}
add_percentiles <- function(stats_list, na.rm = TRUE) {
  
  x <- stats_list$values
  
  c(
    stats_list,
    list(
      p25 = unname(stats::quantile(x, probs = 0.25, na.rm = na.rm)),
      p50 = unname(stats::quantile(x, probs = 0.50, na.rm = na.rm)),
      p75 = unname(stats::quantile(x, probs = 0.75, na.rm = na.rm))
    )
  )
}

stats_extra <- add_percentiles(stats_list = stats)

stats_extra
```

### Listing 9.19. Printing out the `german_cities` dataset, which has two `factor` columns. 

```{r print_german_cities, paged.print=FALSE}
german_cities
```

### Listing 9.20. Printing out the unique values from the state column of `german_cities` (obtained as a vector through `$` indexing). 

```{r print_state_fct}
state_fct <- unique(german_cities$state)
state_fct
```

### Listing 9.21. Getting all the levels for the `state_fct` factor using the `levels()` function.

```{r state_fct_levels}
levels(state_fct)
```

### Listing 9.22. Getting the integer values for the factor levels in `state_fct` with `as.integer()`.

```{r state_fct_as_integer}
as.integer(state_fct)
```

### Listing 9.23. Understanding the frequency of factor levels with the forcats `fct_count()` function.

```{r fct_count, paged.print=FALSE}
fct_count(german_cities$state)
```

### Listing 9.24. Filtering the cities in `german_cities` to those in the state of `Bayern`.

```{r german_cities_bayern, paged.print=FALSE}
german_cities_bayern <-
  german_cities %>%
  filter(state == "Bayern")

german_cities_bayern
```

### Listing 9.25. Dropping unneeded levels from two factor columns of the `german_cities_bayern` tibble.

```{r fct_drop}
german_cities_bayern <- 
  german_cities_bayern %>% 
  mutate(
    name = fct_drop(name),
    state = fct_drop(state)
  )

c(nlevels(german_cities_bayern$name), nlevels(german_cities_bayern$state))
```

### Listing 9.26. Creating a barplot of the most populous German cities in the state of Bayern by using ggplot and its `geom_bar()` function.

```{r barplot_bayern}
ggplot(data = german_cities_bayern, aes(y = name, x = pop_2015)) +
  geom_bar(stat = "identity")
```

### Listing 9.27. Using the `fct_reorder()` function to reorder the factor levels of the `name` variable to match the ordering of the `pop_2015` variable.

```{r fct_reorder}
german_cities_bayern$name <- 
  fct_reorder(german_cities_bayern$name, german_cities_bayern$pop_2015)

levels(german_cities_bayern$name)
```

### Listing 9.26. [re-run].

```{r}
ggplot(data = german_cities_bayern, aes(y = name, x = pop_2015)) +
  geom_bar(stat = "identity")
```

### Listing 9.28. Reversing the order of factor levels in the `name` variable with the `fct_rev()` function.

```{r fct_rev}
german_cities_bayern$name <- fct_rev(german_cities_bayern$name)

levels(german_cities_bayern$name)
```

### Listing 9.26. [re-run].

```{r}
ggplot(data = german_cities_bayern, aes(y = name, x = pop_2015)) +
  geom_bar(stat = "identity")
```

### Listing 9.29. Recoding factor levels in `name` with the `fct_recode()` function.

```{r fct_recode}
german_cities_bayern$name <- 
  fct_recode(
    german_cities_bayern$name,
    "Fuerth" = "Fürth", "Wuerzburg" = "Würzburg",
    "Nuernberg" = "Nürnberg", "Muenchen" = "Munich"
    )

levels(german_cities_bayern$name)
```

### Listing 9.26. [re-run].

```{r}
ggplot(data = german_cities_bayern, aes(y = name, x = pop_2015)) +
  geom_bar(stat = "identity")
```

### Listing 9.30. Creating a bar plot that is based on counts of cities in each `state`, which is mapped to the `y` aesthetic.

```{r barplot_count}
german_cities %>%
  ggplot(aes(y = state)) +
  geom_bar()
```

### Listing 9.31. Improving on the city count bar plot through use of the `fct_infreq()` and `fct_rev()` functions.

```{r fct_infreq}
german_cities %>%
  mutate(state = state %>% fct_infreq() %>% fct_rev()) %>%
  ggplot(aes(y = state)) +
  geom_bar()
```

### Listing 9.32. Using the `fct_inseq()` function to reorder a factor's levels by its numeric values.

```{r fct_inseq}
fctr <- factor(3:8, levels = as.character(c(5, 4, 6, 3, 8, 7)))

fctr %>% levels()

fctr %>% fct_inseq() %>% levels()
```

### Listing 9.33. Further refinement of the city count bar plot by incorporating the `fct_lump_n()` function (just before reversing levels).

```{r fct_lump_n}
german_cities %>%
  mutate(state = state %>% fct_infreq() %>% fct_lump_n(5) %>% fct_rev()) %>%
  ggplot(aes(y = state)) +
  geom_bar()
```
