---
title: "Chapter 5"
output: html_document
---

```{r load_tidyverse_edr_readxl}
library(tidyverse)
library(edr)
library(readxl)
```

## 5.2. Using **readr** to Import Data from CSV Files

### Listing 5.1. Using the `dataset_to_csv()` function (exclusive to the `edr` package and its datasets) to write `us_cities.csv` to the project directory.

```{r}
# dataset_to_csv(dataset = "us_cities")
```

### Listing 5.3. Using the `read_csv()` function to read the `us_cities.csv` to a tibble object.

```{r read_csv, paged.print=FALSE}
city_table <- read_csv(file = "us_cities.csv")

city_table
```

### Listing 5.4. Using the `read_csv()` with `us_cities.csv`, specifying the exact column types in `col_types`.

```{r read_csv_col_types, paged.print=FALSE}
city_table <- read_csv(file = "us_cities.csv", col_types = "cccii")

city_table
```


5.3	Using **readxl** to import Excel data

### Listing 5.5. Using the `create_excel_file()` function to write `us_cities.xlsx` to the project directory.

```{r}
# create_excel_file()
```

### Listing 5.6. Using the `read_excel()` function to read the first sheet from the `us_cities.xlsx` file to a tibble object.

```{r read_excel, paged.print=FALSE}
city_table_x <- read_excel(path = "us_cities.xlsx")

city_table_x
```

### Listing 5.7. Using the `read_excel()` function to read in the second sheet (`us_cities_messy`) from the `us_cities.xlsx` file. The result is not ideal.

```{r read_excel_messy, paged.print=FALSE}
city_table_x_m <- read_excel("us_cities.xlsx", sheet = 2)

city_table_x_m
```

### Listing 5.8. Using the `read_excel()` function with the `us_cities_messy` sheet, take 2. With some options (and some care!) we get a usable tibble.

```{r read_excel_messy_options, paged.print=FALSE}
city_table_x_m <- 
  read_excel(
    "us_cities.xlsx",
    sheet = 2,
    skip = 2,
    col_types = c(rep("text", 3), rep("numeric", 2), rep("skip", 2)),
    na = "N/A")

city_table_x_m
```

### Listing 5.9. The column names (after importing from Excel) can be easily changed using dplyr’s `rename()` function.

```{r rename_columns, paged.print=FALSE}
city_table_x_m <- 
  city_table_x_m %>%
  rename(
    state_id = `state id`,
    state_name = `state name`,
    pop_urb = `urban population`,
    pop_mun = `municipal population`
  )

city_table_x_m
```
