---
title: "Chapter 6"
output: html_document
---

## 6.2. Using **lubridate** to Parse Date and Date-Time Strings

### Listing 6.1. Loading the **tidyverse**, **lubridate**, and **edr** packages

```{r load_tidyverse_lubridate_edr}
library(tidyverse)
library(lubridate)
library(edr)
```

### Listing 6.2. Examples of typical date formats with different orderings of year, month, and day (all for the same date)

```{r text_dates}
ymd_chr <- "19950615"
mdy_chr <- "06/15/1995"
dmy_chr <- "15-06-1995"
```

### Listing 6.3. Converting text-based dates to R `Date` objects.

```{r to_date_objects}
date_1 <- ymd(ymd_chr)
date_2 <- mdy(mdy_chr)
date_3 <- dmy(dmy_chr)

dates <- c(date_1, date_2, date_3)

dates
```

### Listing 6.4. Checking that we have a `Date` vector.

```{r check_class_date}
class(dates)
```

### Listing 6.5. Creating a `Date` object from a date string in ISO notation.

```{r as_date_function}
as_date("19950615")
```

### Listing 6.6. Typical date formats with an added time component.

```{r text_date_times}  
ymd_hm_chr <- "19950615 0800"
mdy_hm_chr <- "06/15/1995 08:00"
dmy_hm_chr <- "15-06-1995 08:00"
```

### Listing 6.7. Converting text-based date-times to R `POSIXct` objects.

```{r to_datetime_objects}
datetime_1 <- mdy_hm(mdy_hm_chr)
datetime_2 <- dmy_hm(dmy_hm_chr)
datetime_3 <- ymd_hm(ymd_hm_chr)

datetimes <- c(datetime_1, datetime_2, datetime_3)

datetimes
```

### Listing 6.8. Checking that we have a `POSIXct` vector.

```{r check_class_posixct}
class(datetimes)
```

## 6.3. Transforming Dates and Times in Tabular Data

### Listing 6.9. The `nycweather` dataset, printed to the console.

```{r print_nycweather, paged.print=FALSE}
nycweather
```

### Listing 6.10. Plotting the dataset (as is) with **ggplot**.

```{r plot_nycweather_asis}
ggplot(data = nycweather) +
  geom_point(aes(x = time, y = temp))
```

### Listing 6.11. Plotting the beginning and end of `nycweather`.

```{r plot_nycweather_start_end}
nycweather_start_end <-
  dplyr::bind_rows(
    head(nycweather, 100),
    tail(nycweather, 100)
  )

ggplot(data = nycweather_start_end) +
  geom_point(aes(x = time, y = temp))
```

### Listing 6.12. Transforming character times to `POSIXct` date-times.

```{r make_datetime_column, paged.print=FALSE}
nycweather_a <- 
  nycweather %>%
  mutate(time = as_datetime(time))

nycweather_a
```

### Listing 6.13. Using a sample of our dataset for quicker plotting.

```{r downsample_for_plotting, paged.print=FALSE}
nycweather_a_samp <-
  nycweather_a %>%
  sample_n(size = 1000) %>%
  filter(!is.na(temp))

nycweather_a_samp
```

### Listing 6.14. Plotting a sample of the revised dataset.

```{r plot_with_sampled_data}
ggplot(data = nycweather_a_samp) +
  geom_point(aes(x = time, y = temp))
```

### Listing 6.15. Reducing our dataset to get one measurement per hour.

```{r one_per_hour, paged.print=FALSE}
nycweather_b <- 
  nycweather_a %>%
  filter(!is.na(temp)) %>%
  mutate(
    date = date(time),
    hour = hour(time)
  ) %>%
  group_by(date, hour) %>%
  slice(1) %>%
  ungroup()

nycweather_b
```

### Listing 6.16. Plotting our hourly temperatures as a line.

```{r plot_hourly_data_line}
ggplot(data = nycweather_b) +
  geom_line(aes(x = time, y = temp))
```

### Listing 6.17. Summarizing our data to get daily minimum and maximum temperatures.

```{r get_daily_min_max, paged.print=FALSE}
nycweather_c <-
  nycweather_b %>%
  group_by(date) %>%
  summarize(
    min  = min(temp),
    max  = max(temp)
  )
  
nycweather_c
```

### Listing 6.18. Tidying the table to get one temperature value per row.

```{r tidy_table, paged.print=FALSE}
nycweather_d <- 
  nycweather_c %>%
  pivot_longer(
    c(min, max),
    names_to = "type",
    values_to = "temp"
  ) %>%
  arrange(date)

nycweather_d
```

### Listing 6.19. Plotting our daily minimum and maximum temperatures.

```{r plot_daily_min_max_lines}
ggplot(data = nycweather_d) +
  geom_line(aes(x = date, y = temp, color = type)) +
  scale_color_manual(values = c("red", "blue")) +
  scale_x_date(date_breaks = "2 months", date_minor_breaks = "1 month") + 
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.position = "top",
    legend.justification = "left"
  ) + 
  labs(
    title = "Daily Minimum and Maximum Temperatures",
    subtitle = "2010 Data, Measured at JFK Airport",
    x = "Date",
    y = "Temperature (degrees C)"
  )
```

