---
title: "Chapter 3"
output: html_document
---

```{r load_tidyverse_edr}
library(tidyverse)
library(edr)
```

## 3.1. Using **ggplot** to Create Plots

### Listing 3.1. Code to make our first ggplot.

```{r gg_empty}
ggplot(data = dmd, mapping = aes(x = carats, y = price))
```

### Listing 3.2. Using `geom_point()` adds a layer of points.

```{r dmd_carats_price}
ggplot(data = dmd, mapping = aes(x = carats, y = price)) +
  geom_point()
```

### Listing 3.3. Using a different set of y values.

```{r dmd_carats_depth}
ggplot(data = dmd, mapping = aes(x = carats, y = depth)) +
  geom_point()
```

### Listing 3.4. Using a different set of y values.

```{r dmd_shape_for_clarity}
ggplot(dmd, mapping = aes(x = carats, y = price, shape = clarity)) +
  geom_point()
```

### Listing 3.5. Using the `alpha` argument in `geom_point()`.

```{r dmd_carats_price_alpha}
ggplot(dmd, mapping = aes(x = carats, y = price, shape = clarity)) +
  geom_point(alpha = 0.25)
```

### Listing 3.6. Setting a fixed `color` inside of `geom_point()`.

```{r dmd_all_gray50_points}
ggplot(dmd, mapping = aes(x = carats, y = price)) +
  geom_point(color = "gray50")
```

### Listing 3.7. Using color and shape aesthetics.

```{r dmd_aesthetics_1}
ggplot(dmd, aes(x = carats, y = price)) +
  geom_point(mapping = aes(color = cut, shape = clarity))
```

### Listing 3.8. Using the size aesthetic and a fixed alpha.

```{r dmd_aesthetics_2}
ggplot(dmd, aes(x = carats, y = price)) +
  geom_point(mapping = aes(size = depth), alpha = 0.05)
```

### Listing 3.9. Supplying fixed values for color, fill, and shape.

```{r dmd_aesthetics_3}
ggplot(dmd, aes(x = carats, y = price)) +
  geom_point(color = "gray50", fill = "#AAAFEF", shape = 23)
```

### Listing 3.10. Applying the color aesthetic to the depth variable.

```{r dmd_aesthetics_4}
ggplot(dmd, aes(x = carats, y = price)) +
  geom_point(mapping = aes(color = depth, shape = clarity))
```

### Listing 3.11. Using an additional statement with `facet_wrap()` gives us a faceted plot.

```{r dmd_facet_clarity}
ggplot(dmd, aes(x = carats, y = price)) +
  geom_point() +
  facet_wrap(facets = vars(clarity))
```

### Listing 3.12. Faceting by two variables: `cut` and `clarity`.

```{r dmd_facet_cut_clarity}
ggplot(dmd, aes(x = carats, y = price)) +
  geom_point() +
  facet_wrap(facets = vars(cut, clarity))
```

### Listing 3.13. Using the labeller function `label_both` to create informative labels for facets.

```{r dmd_facet_cut_clarity_labeller}
ggplot(dmd, aes(x = carats, y = price)) +
  geom_point() +
  facet_wrap(facets = vars(cut, clarity), labeller = label_both)
```

### Listing 3.14. We can specify the total number of rows of plot panels with the `nrow` argument.

```{r dmd_facet_cut_clarity_wide}
ggplot(dmd, aes(x = carats, y = price)) +
  geom_point() +
  facet_wrap(
    facets = vars(cut, clarity),
    nrow = 1, labeller = label_both
  )
```

### Listing 3.15. Using `facet_grid()` provides a slightly different visualization of the faceted plot panels.

```{r dmd_facet_cut_clarity_grid}
ggplot(dmd, aes(x = carats, y = price)) +
  geom_point(alpha = 0.2) +
  facet_grid(rows = vars(cut), cols = vars(clarity), labeller = label_both)
```

### Listing 3.16. The `labs()` function gives us the opportunity to provide our own labels for different plot elements.

```{r plot_new_labs}
ggplot(dmd, mapping = aes(x = carats, y = price)) +
  geom_point(mapping = aes(shape = clarity)) +
  labs(
    x = "Weight of the Diamond (carats)",
    y = "Price (USD)",
    shape = "Diamond Clarity"
  )
```

### Listing 3.17. We can specify the plot’s title, subtitle, and caption inside `labs()` as well.

```{r plot_labs_titles}
ggplot(dmd, mapping = aes(x = carats, y = price)) +
  geom_point(mapping = aes(shape = clarity)) +
  labs(
    title = "The Relationship Between Diamond Weight on Price",
    subtitle = "Quality of diamond clarity is indicated by shape",
    caption = "Data taken from the `dmd` dataset",
    x = "Weight of the Diamond (carats)", y = "Price (USD)",
    color = "Diamond Cut", shape = "Diamond Clarity"
  )
```

### Listing 3.18. Using `legend.position` argument of `theme()` to put the legend to the right of the plot.

```{r plot_legend_right}
ggplot(dmd, aes(x = carats, y = price)) +
  geom_point(aes(shape = clarity)) +
  labs(shape = "Clarity") +
  theme(legend.position = "right")
```

### Listing 3.19. Using `legend.position` argument of `theme()` to put the legend below the plot.

```{r plot_legend_bottom}
ggplot(dmd, aes(x = carats, y = price)) +
  geom_point(aes(shape = clarity)) +
  labs(shape = "Clarity") +
  theme(legend.position = "bottom")
```

### Listing 3.20. Using `legend.justification` argument of `theme()` to justify the legend toward the top of the visualization.

```{r plot_legend_just_top}
ggplot(dmd, aes(x = carats, y = price)) +
  geom_point(aes(shape = clarity)) +
  labs(shape = "Clarity") +
  theme(legend.justification = "top")
```

### Listing 3.21. Using `legend.position` argument of `theme()` to remove the legend entirely.

```{r plot_no_legend}
ggplot(dmd, aes(x = carats, y = price)) +
  geom_point(aes(shape = clarity)) +
  labs(shape = "Clarity") +
  theme(legend.position = "none")
```

### Listing 3.22. Modifying `dmd` to obtain two new columns: `cpc` and `price_class`.

```{r dmd_modified_dplyr_1}
dmd_mod <- 
  dmd %>%
  mutate(cpc = price / carats) %>%
  mutate(price_class = ifelse(cpc >= 3460, "Above Median", "Below Median"))
```

### Listing 3.23. A third and final `mutate()` statement to add the `quality` column to our modified dataset (`dmd_mod`).

```{r dmd_modified_dplyr_2}
dmd_mod <- 
  dmd %>%
  mutate(cpc = price / carats) %>%
  mutate(price_class = ifelse(cpc >= 3460, "Above Median", "Below Median")) %>%
  mutate(quality = ifelse(
    cut == "The Best" & clarity == "The Best",
    "Top Drawer", "The Rest")
  )
```

### Listing 3.24. Modifying `dmd` to add three new columns, and, plotting `dmd_mod` with the new variables.

```{r dmd_modified_dplyr_ggplot}
dmd_mod <-
  dmd %>%
  mutate(cpc = price / carats) %>%
  mutate(price_class = ifelse(cpc >= 3460, "Above Median", "Below Median")) %>%
  mutate(quality = ifelse(
    cut == "The Best" & clarity == "The Best", "Top Drawer", "The Rest")
  )

ggplot(dmd_mod, aes(x = carats, y = price)) +
  geom_point() +
  facet_wrap(
    facets = vars(price_class, quality),
    labeller = label_both
  ) +
  labs(x = "Carats", y = "Price")
```

