---
title: "Chapter 12"
output: html_document
---

```{r load_packages}
library(tidyverse)
library(edr)
library(DT)
library(gt)
```

## 12.1. Making Presentation Tables

### Listing 12.1. A simple **DT** table using the `pitchfork` dataset.

```{r dt_pitchfork_1}
pitchfork_1000 <- pitchfork %>% dplyr::slice_head(n = 1000)

datatable(pitchfork_1000)
```

### Listing 12.2. Using dplyr's `mutate()` function to help us transform bare links into HTML links.

```{r dt_pitchfork_links, paged.print=FALSE}
pitchfork_1000 <- 
  pitchfork_1000 %>%
  mutate(link = paste0("<a href='", link ,"'>review</a>"))

pitchfork_1000 %>% select(link)
```

### Listing 12.3. Making the column of links functional in a **DT** table by removing the `escape` condition from column `8`.

```{r dt_pitchfork_2}
pitchfork_1000 %>% datatable(escape = -8)
```

### Listing 12.4. Changing the table appearance of the **DT** table with the `style` and `class` parameters, and, by removing the column with row names.

```{r dt_pitchfork_3}
pitchfork_1000 %>%
  datatable(
    escape = -7,
    style = "bootstrap4",
    class = "table-bordered",
    rownames = FALSE
  )
```

### Listing 12.5. The column names of the **DT** table can be modified using the `colnames` argument.

```{r dt_pitchfork_4}
pitchfork_1000 %>%
  datatable(
    escape = -7,
    style = "bootstrap4",
    class = "table-bordered",
    rownames = FALSE,
    colnames = stringr::str_to_title(colnames(pitchfork_1000))
  )
```

### Listing 12.6. Adding filtering functionality to the **DT** table (UI applied to the `top` of the table).

```{r dt_pitchfork_5}
pitchfork_1000 %>%
  datatable(
    escape = -7,
    style = "bootstrap4",
    class = "table-bordered",
    rownames = FALSE,
    colnames = stringr::str_to_title(colnames(pitchfork_1000)),
    filter = "top"
  )
```

### Listing 12.7. Adding useful buttons for exporting the table data to various files, copying the data, and printing.

```{r dt_pitchfork_6}
pitchfork_1000 %>%
  datatable(
    escape = -7,
    style = "bootstrap4",
    class = "table-bordered",
    rownames = FALSE,
    colnames = stringr::str_to_title(colnames(pitchfork_1000)),
    filter = "top",
    extensions = "Buttons",
    options = list(
      buttons = c("csv", "excel", "pdf", "copy", "print"),
      dom = "Bfrtip"
    )
  )
```


### Listing 12.8

```{r gt_exibble_1}
gt_tbl_1 <- exibble %>% gt()

gt_tbl_1
```

### Listing 12.9

```{r gt_exibble_2}
gt_tbl_2 <-
  gt_tbl_1 %>%
  tab_header(
    title = "The exibble dataset",
    subtitle = "(It's available in the gt package.)"
  )

gt_tbl_2
```


### Listing 12.10

```{r gt_exibble_3}
gt_tbl_3 <-
  gt_tbl_2 %>%
  tab_source_note("The gt package is available in CRAN.") %>%
  tab_source_note("This example is provided by the edr book.")

gt_tbl_3
```


### Listing 12.11

```{r gt_exibble_4}
gt_tbl_4 <-
  gt_tbl_3 %>%
  tab_footnote(
    footnote = "This coconut contains a tape and half a necklace.",
    locations = cells_body(columns = vars(char), rows = 3)
  )

gt_tbl_4
```

### Listing 12.12

```{r gt_exibble_5}
gt_tbl_5 <-
  gt_tbl_4 %>%
  tab_footnote(
    footnote = "This column contains alphabetical fruit.",
    locations = cells_column_labels(columns = vars(char))
  )

gt_tbl_5
```

### Listing 12.13. Making the `oceaniapops` table from the (much larger) `countrypops` table.

```{r oceaniapops_countrypops, paged.print=FALSE}
Australasia <- c("AU", "NZ")
Melanesia <- c("NC", "PG", "SB", "VU")
Micronesia <- c("FM", "GU", "KI", "MH", "MP", "NR", "PW")
Polynesia <- c("PF", "WS", "TO", "TV")

oceaniapops <- 
  countrypops %>%
  dplyr::filter(country_code_2 %in% c(
    Australasia, Melanesia, Micronesia, Polynesia)
  ) %>%
  dplyr::filter(year %in% c(1995, 2005, 2015)) %>%
  dplyr::mutate(region = case_when(
    country_code_2 %in% Australasia ~ "Australasia",
    country_code_2 %in% Melanesia ~ "Melanesia",
    country_code_2 %in% Micronesia ~ "Micronesia",
    country_code_2 %in% Polynesia ~ "Polynesia"
  )) %>%
  tidyr::pivot_wider(names_from = year, values_from = population) %>%
  dplyr::arrange(region, desc(`2015`)) %>%
  dplyr::select(-starts_with("country_code"))

oceaniapops
```


### Listing 12.14. Making a very simple **gt** simple with the `oceaniapops` tibble.

```{r gt_oceaniapops_0}
oceaniapops %>% gt()
```

### Listing 12.15. Taking a different approach in making the **gt** table by defining a stub and row group labels.

```{r gt_oceaniapops_1}
oceaniapops_1 <- 
  oceaniapops %>%
  gt(
    rowname_col = "country_name",
    groupname_col = "region"
  )

oceaniapops_1
```

### Listing 12.16. Adding more structure to the **gt** table by placing the spanner label "Total Population" above the column labels.

```{r gt_oceaniapops_2}
oceaniapops_2 <- 
  oceaniapops_1 %>%
  tab_spanner(
    label = "Total Population",
    columns = vars(`1995`, `2005`, `2015`)
  )

oceaniapops_2
```

### Listing 12.17. We can make the population values much to easier to read with decimal separators by use of the `fmt_number()` function.

```{r gt_oceaniapops_3}
oceaniapops_3 <- 
  oceaniapops_2 %>%
  fmt_number(
    columns = vars(`1995`, `2005`, `2015`),
    decimals = 0
  ) %>%
  cols_align(align = "right")

oceaniapops_3
```

### Listing 12.18. The column widths can be finely adjusted with `cols_width()`, and, a heading is always a good idea.

```{r oceaniapops_4}
oceaniapops_4 <- 
  oceaniapops_3 %>%
  cols_width(
    1 ~ px(250),
    everything() ~ px(125)
  ) %>%
  tab_header(
    title = md("Populations of Countries in Oceania During **1995**, **2005**, and **2015**"),
    subtitle = "Combined land area of 3,291,903 sq mi and a population of over 41 million."
  )

oceaniapops_4
```

### Listing 12.19. Footnotes and sources can be added to a **gt** with the `tab_footnote()` and `tab_source_note()` functions.

```{r oceaniapops_5}
oceaniapops_5 <- 
  oceaniapops_4 %>%
  tab_footnote(
    footnote = "United States.",
    locations = cells_stub(rows = starts_with(c("No", "Gu")))
  ) %>%
  tab_footnote(
    footnote = "France.",
    locations = cells_stub(rows = starts_with(c("New Cal", "French")))
  ) %>%
  tab_source_note(
    source_note = md(
      paste0(
        "Population data obtained from [World Bank Open Data]",
        "(https://data.worldbank.org/indicator/SP.POP.TOTL)."
      )
    )
  )

oceaniapops_5
```

### Listing 12.20. Text can be boldened with combination of the `tab_style()` and `cell_text()` functions; we can do this at multiple locations at once.

```{r oceaniapops_6}
oceaniapops_6 <- 
  oceaniapops_5 %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(
      cells_column_labels(columns = TRUE),
      cells_column_spanners(spanners = TRUE),
      cells_row_groups(groups = TRUE)
    )
  )

oceaniapops_6
```

### Listing 12.21. Summary rows can be generated on a per-group basis (with `summary_rows()`), and, for all rows in the table (with `grand_summary_rows()`).

```{r}
oceaniapops_7 <- 
  oceaniapops_6 %>%
  summary_rows(
    groups = TRUE,
    columns = TRUE,
    fns = list(TOTAL = ~ sum(., na.rm = FALSE)),
    formatter = fmt_number,
    decimals = 0
  ) %>%
  grand_summary_rows(
    columns = TRUE,
    fns = list(`GRAND TOTAL` = ~ sum(., na.rm = FALSE)),
    formatter = fmt_number,
    decimals = 0
  ) %>%
  tab_options(data_row.padding = px(4))
  
oceaniapops_7
```

