---
title: "Chapter 8"
output: html_document
---

```{r load_packages}
library(tidyverse)
library(edr)
```

### Listing 8.1. Assigning a character string to `one_word`.

```{r one_word}
one_word <- "hello"
one_word
```

### Listing 8.2. Creating a character vector with three elements.

```{r three_words}
three_words <- c("one", "two", "three")
```

### Listing 8.3. Combining two character vectors with the `c()` function.

```{r words_and_days}
days <- c("monday", "tuesday", "wednesday")

words_and_days <- c(three_words, days)

words_and_days
```

### Listing 8.4. Combining a character vector with a numeric vector.

```{r words_and_numbers}
three_numbers <- c(1, 2, 3)

words_and_numbers <- c(three_words, three_numbers)

words_and_numbers
```

### Listing 8.5. How combining elements of different types results in coercion.

```{r coercion_to_character}
c(7, 1.2, TRUE, 6.4/2, FALSE, "hello")
```

### Listing 8.6. Creating a data frame with two columns.

```{r data_frame_w_factors, paged.print=FALSE}
data_frame <- data.frame(a = three_numbers, b = three_words)
data_frame
```

### Listing 8.7. Using the `stringsAsFactors` option with `data.frame()`.

```{r data_frame_no_factors, paged.print=FALSE}
data_frame_2 <- 
  data.frame(
    a = three_numbers,
    b = three_words,
    stringsAsFactors = FALSE
  )

str(data_frame_2)
```


### Listing 8.8. Using the `formatC()` function to get very particular output character strings for three decimal numbers.

```{r formatC_decimals_2dig}
decimal_numbers <- c(3.825, 1.01, 15.7758)

formatC(
  decimal_numbers,
  digits = 3,
  format = "f",
  decimal.mark = ".",
  drop0trailing = FALSE
)
```

### Listing 8.9. Using `formatC()` in a `mutate()` call to format sequential numeric values to index strings that are left-padded with zeros (to a specified width).

```{r formatC_mutate, paged.print=FALSE}
combined_tbl <-
  dplyr::tibble(index = 1:1200, text = rep(LETTERS[1:12], 100)) %>%
  dplyr::mutate(index = formatC(
    index,
    digits = 4,
    format = "d",
    flag = "0"
  )) %>%
  tidyr::unite(col = "combined", index, text) %>%
  dplyr::arrange(desc(combined))

combined_tbl
```

### Listing 8.10. Using the `toupper()` and `tolower()` functions to transform characters to upper case and lower case.

```{r toupper_tolower}
mixed_cases <- c("One", "two", "Three", "FOUR")

c(
  toupper(mixed_cases[1:2]),
  tolower(mixed_cases[3:4])
)
```

### Listing 8.11. Using `abbreviate()` to make unique abbreviations for longer strings.

```{r abbreviate_terms}
terms <- c("location", "site", "area", "arena", "locale", "site")

abbreviate(terms, minlength = 3)
```

### Listing 8.12. Using `substr()` to truncate very long strings to a specific length.

```{r substr_resto_reviews}
resto_reviews_trunc <- resto_reviews %>% substr(1, 80)

resto_reviews_trunc[1:5]
```

### Listing 8.13. Using `trimws()` to remove any leading and trailing whitespace in our truncated strings.

```{r trimws_resto_reviews_trunc}
resto_reviews_trunc <- resto_reviews_trunc %>% trimws()

resto_reviews_trunc[1:5]
```

### Listing 8.14. Using `strtrim()` to truncate strings, much like `substr()`.

```{r strtrim_resto_reviews}
all.equal(resto_reviews %>% substr(1, 80), resto_reviews %>% strtrim(80))
```

### Listing 8.15. Opening up the RegExplain gadget for experimenting with regular expressions.

```{r open_regexplain}
# edr_regexplain_gadget()
```

### Listing 8.16. Using `str_match()` to determine the first match of the regex pattern on the text.

```{r str_match}
fox_dog <- "The quick brown fox jumps over the lazy dog."

str_match(fox_dog, "....")
```

### Listing 8.17. Using `str_match_all()` to get all of the matches from the text.

```{r str_match_all}
str_match_all(fox_dog, ".o.")
```

### Listing 8.18. A regex example that uses character sets, character classes, and negation.

```{r character_sets_classes}
ihavefruit <- "Fruit? I've got a lime, 3 figs, and 5 plums."

str_match_all(ihavefruit, "[\\d\\w][^A-Z]\\w\\w\\w\\w\\S")
```

### Listing 8.19. Using the `?` quantifier to denote an optional piece of text.

```{r character_sets_classes}
str_match_all(ihavefruit, "[\\d\\w]\\s\\w\\w\\w\\w?s?\\S")
```

### Listing 8.20. Using anchors to bind the search pattern to the start and/or end of the string.

```{r anchors}
str_match(ihavefruit, "^\\w*")
str_match(ihavefruit, "\\w*\\.?$")
str_match(ihavefruit, "^[[:print:]]*$")
```

### Listing 8.21. Using alternation (with `|`) inside of parentheses (which is a capturing group).

```{r grouping_capturing_alternation}
str_match(ihavefruit, "(figs?|plums?|limes?)")
```

### Listing 8.22. Backreferences are used to compose a new string in a particular order with the captured text.

```{r str_replace_all}
ihavefruit %>%
  str_replace_all(
    pattern = ".*(limes?).*(figs?).*(plums?).*",
    replacement = "\\3 \\2 \\1"
  ) %>%
  str_replace_all(pattern = "s\\b", replacement = "") %>%
  str_split(pattern = " ") %>%
  unlist()
```

### Listing 8.23. Extracting words from a string and putting them into a vector with `str_extract_all()`.

```{r str_extract_all}
ihavefruit %>%
  str_extract_all(pattern = "(limes?|figs?|plums?)") %>%
  unlist()
```

### Listing 8.24. Improving the `pattern` with the word boundary metacharacter.

```{r str_extract_all_word_boundary}
ihavefruit_2 <- "Fruit? Yep, I've got a lime, 3 figs, and 5 plums. All covered in slime."

ihavefruit %>%
  str_extract_all(pattern = "(\\blimes?\\b|\\bfigs?\\b|\\bplums?\\b)") %>%
  unlist()
```

### Listing 8.25. Using the `str_detect()` function to see whether there are any matches of our pattern in the `resto_reviews` vector.

```{r str_detect}
str_detect(string = resto_reviews, pattern = "yum\\w*?") %>% 
  table()
```

### Listing 8.26. Using the logical vector from `str_detect()` to inspect those reviews that matched the regex.

```{r str_detect_see_matches}
matches <- str_detect(string = resto_reviews, pattern = "yum\\w*?")

resto_reviews[matches]
```

### Listing 8.27. Getting all matches of dollar values and eventually resolving to a single numeric vector.

```{r str_extract_all_str_remove}
str_extract_all(string = resto_reviews, pattern = "\\$\\d*") %>%
  unlist() %>%
  str_remove("\\$") %>%
  as.numeric()
```


