---
title: "Chapter 4"
output: html_document
---

```{r load_tidyverse_edr}
library(tidyverse)
library(edr)
```

## 4.2. Using **tidyr** to Tidy Our Tables

### Listing 4.1. The `winniweather` dataset, printed to the console.

```{r winniweather_print, paged.print=FALSE}
winniweather
```

### Listing 4.2. How to make data `pivot_longer()` and thus make it more tidy.

```{r winni_pivot_longer, paged.print=FALSE}
winni_mod <- 
  winniweather %>%
  pivot_longer(
    starts_with("temp"),
    names_to = "hour",
    values_to = "temp"
  )

winni_mod
```

### Listing 4.3. Modifying values in the `hour` column with `mutate()`.

```{r mutate_case_when, paged.print=FALSE}
winni_mod <- 
  winni_mod %>%
  mutate(hour = case_when(
    hour == "temp00_00" ~ 0L,
    hour == "temp06_00" ~ 6L,
    hour == "temp12_00" ~ 12L,
    hour == "temp18_00" ~ 18L
  ))

winni_mod
```

### Listing 4.4. Separating the `yearmonth` column into `year` and `month` columns.

```{r separate, paged.print=FALSE}
winni_tidy <- 
  winni_mod %>%
  separate(
    col = yearmonth,
    into = c("year", "month"),
    sep = "-",
    convert = TRUE
  )

winni_tidy
```

### Listing 4.5. Using `arrange()` to put the observations in the correct order.

```{r weather_arrange, paged.print=FALSE}
winni_tidy <- 
  winni_tidy %>%
  arrange(year, month, day, hour)

winni_tidy
```

### Listing 4.6. A date-time column is necessary for ggplot time-series plots.

```{r winni_iso_date, paged.print=FALSE}
winni_tidy <- 
  winni_tidy %>%
  mutate(iso_date = ISOdate(year, month, day, hour))

winni_tidy
```

### Listing 4.7. A scatterplot of all observations in `winni_tidy` using ggplot.

```{r winni_tidy_ggplot_1}
ggplot(data = winni_tidy) +
  geom_point(aes(x = iso_date, y = temp))
```

### Listing 4.8. Replacement of missing values with the `mutate()` and `na_if()` functions.

```{r winni_na_if, paged.print=FALSE}
winni_tidy <-
  winni_tidy %>%
  mutate(temp = na_if(temp, 9999))
```

### Listing 4.9. Inspecting the data in a column to verify that the mutation had occurred.

```{r winni_temp_summary, paged.print=FALSE}
winni_tidy %>% 
  select(temp) %>%
  summary()
```

### Listing 4.10. An updated scatterplot of all observations in `winni_tidy` after encoding a missing value.

```{r winni_tidy_ggplot_2}
ggplot(data = winni_tidy) +
  geom_point(aes(x = iso_date, y = temp))
```

### Listing 4.11. Faceting the temperatures in `winni_tidy` by the time of day.

```{r winni_tidy_ggplot_3}
ggplot(data = winni_tidy) +
  geom_point(aes(x = iso_date, y = temp)) +
  facet_wrap(vars(hour), labeller = label_both)
```

