--- 
title: Rolling Up Data
---

```{r setup, include=FALSE, echo=FALSE}
library(dspatterns)
library(palmerpenguins)
```

# Rolling Up Data

Notes: 

- focusing on explaining data by pulling out data bits from objects (e.g., summary stats)
- feature gtsummary()
- explicitly numeric variables - NOT ones that you want to count essentially
- https://apreshill.github.io/data-vis-labs-2018/04-distributions.html
- this chapter might be slightly opinionated about which stats to use when? (i.e., median is always useful)
- the idea being people may simply report means b/c they are easy to calculate, and easy to explain. But this chapter is going to take a different view: there is nothing harder about calculating other stats, or explaining them with the right tools.

## This chapter's pattern

- Explore: summary / describe / skimr /  gtsummary::tbl_summary() [default is median yay] / looking at & fixing missingness
- Understand: tables (gtsummary + gt) + viz (distributions - boxplots / beeswarms / stat_*) / layer_data() to understand where the data is in a ggplot layer
- Explain: pull / pluck + combine with glue for inline code? (need ways to extract individual cells from gt/gtsummary) https://www.tjmahr.com/lists-knitr-secret-weapon/
- Share (Collaborate/Communicate): use pointblank; gtsummary with better formatting (Median (IQR) )

## test


```{r}
penguins %>% select(where(is.numeric))
```


