--- 
knit: quarto render
---

# Careful Counting Patterns {#sec:careful-counting-patterns}

```{r setup, include=FALSE, echo=FALSE}
library(dspatterns)
library(tidyverse)
library(bakeoff)
```

```{css, echo=FALSE}

.sourceCode {
  font-size: small;
}

.cell-output-stdout {
  font-size: small;
}

.gt_table {
  font-size: small;
}
```


## Introduction

Notes: 

- Use the bakeoff package (diamonds has been seen before)
  - user the bakers datasets
  - how many bakers...
  
This is really good for interactive use. Gives you the rows and columns plus some of row content.
```{r how-many-bakers-glimpse}
glimpse(bakers)
```

Use the chunk option `paged.print=TRUE` to print tables as an HTML table that has controls for pagination.

```{r baker-paged, paged.print=TRUE}
bakers
```

Use the chunk option `paged.print=FALSE` to give us the 'console' version of the table.

```{r bakers-non-paged, paged.print=FALSE}
bakers
```


This is good for sharing a preview of the dataset. Use gt's `gt_preview()` function when you want to share in a static document.

```{r bakers-gt-preview}
gt::gt_preview(bakers)
```



```{r}
bakers %>% nrow()
```

But are the contestants unique across all the rows? (Did somebody drop out and come back?)

```{r, paged.print=FALSE}
bakers %>% distinct(baker_full)
```

QLC: how would you get a count of distinct rows in the `bakers` table.
```{r}
bakers %>% n_distinct() %>% nrow()
```
- most common first name `baker_first`?

```{r}
bakers %>% distinct(baker_first)
```

`107` is less than the `120` row count we've seen earlier. Let's instead `count()` by first name (`baker_first`)

```{r}
bakers %>% count(baker_first)
```


This is more apparent when sorting with the `sort` option in `count()`.

```{r}
bakers %>% count(baker_first, sort = TRUE)
```

Now we see that the most common first name is `Kate`.

This is useful but a bit confusing because of the sorting (only descending by `name_total`).

```{r}
bakers %>% 
  distinct(series, baker_first) %>%
  add_count(baker_first, sort = TRUE, name = "name_total")
```

   - overall?

Let's do a count of baker by series (were there any series where two people shared the same name?)


```{r}
bakers %>% count(series, baker_first, sort = TRUE)
```
Seems like no? All `n` values are `1` and the total count here is the same as before (`120`). 

```{r}
bakers %>% count(series)
```
This follows the original order of the table (series increasing). Interestingly, the first series started with only 10 but it became at least bakers dozen's worth later on. 

```{r}
bakers_by_series <- bakers %>% count(series)
```

In R Markdown, we often want to use the things we calculated inside the text. We can do that with `r knitr::inline_expr("min(bakers_by_series$n)")`
 
    - per series?
    - per hometown?
    
EXERCISE: Can you figure out if any bakers are from the same hometown? How about from the same hometown and the same series?

```{r}
bakers %>% count(hometown, series, sort = TRUE)
```
    
    - per episode / per series?
    
```{r}
episodes %>% count(series, episode)
```



- This chapter really ought to be an intro vis / data inspection/exploration section with multiple packages
- introduce functions:
  - dplyr::glimpse()
  - dplyr::distinct() (getting unique rows by selection of columns)
  - dplyr::count() (counting distinct values by rows; count by series, use built-in sort feature)
  - geom_boxplot()
- use some code from course material that uses bakeoff
  - links: https://alison.netlify.app/rls-plot-twist/#19, https://alison.netlify.app/rls-plot-twist/#19 (counting slides)
- start off with histogram and create this plot: https://alison.netlify.app/rls-plot-twist/#19
- histogram of ages of bakers + other interesting plots
- goals: making exploratory bar charts and histograms (useful for initial understanding of the data)

## Patterns

Explore: exploring a brand new dataset with dplyr and ggplot

Understand: from exploratory plots (confirms or disproves your assumptions about the data) gain an intuition and develop that ability

Explain:

Collaborate:

Communicate:

## Project Ideas

(should have a fully working set of Rmd templates, presented in the form of a project template - this can be prepped for the reader with a function call)

The moma dataset (https://apreshill.github.io/data-vis-labs-2018/02-moma.html) would be great for a project
- explore using the lisa palette from paletteer





### Intro


### Data


### Goals


### Instructions

 - instructions on using project wizard


### Packages Needed

 - installation details (script with `library()` calls)
 - possibly use: https://github.com/rstudio4edu/rmd4edu/blob/master/R/lesson.R


