# Writing

```{r setup, include=FALSE, echo=FALSE}
library(dspatterns)
library(gt)
library(dplyr)
```


Everything that is not code in your `.qmd` is called narrative text, and can be styled with Markdown. Markdown is a plain text formatting syntax, which means everything you see in the file itself can be typed with a normal keyboard. Below is an example:


## Add structure

You can use Markdown to organize your `.qmd` file with headings. Headings are useful for organizing your document into sections, and with Markdown formatting you can have up to 6 types/sizes of headers. Use a hash symbol like `#` before the label to create a new header.

drop-down IDE navigator

outline pane

## Add formatting

All narrative text in your `.qmd` file can be formatted to add **bold**, *italics*, bulletted lists, numbered lists, superscripts<sup>1</sup>, and subscripts<sub>1</sub>.

## Insert elements

do side by side here

<https://rstudio.github.io/visual-markdown-editing/#/>

here add a bibliography, a static image, equation,

<https://rmd4medicine.netlify.app/book/_book/add-text-with-markdown.html>

this is markdown

## What is Markdown?

-   Markdown is a particular type of markup language.

-   Markup languages are designed produce documents from plain text.

## Quick reference

```{r echo=FALSE}
tribble(
  ~a,                                         ~b,
  "`Plain text`",                             "Plain text",
  "`*italics* and **bold**`",                 "*italics* and **bold**",
  "`verbatim code`",                          "`verbatim code`",
  "`~~strikethrough~~`",
  "<span style='text-decoration:line-through;'>strikethrough</span>",
  "`subscript~2~ / superscript^2^`",          "subscript<sub>2</sub> / superscript<sup>2</sup>",
  "`[link](https://www.rstudio.com)`",        "[link](https://www.rstudio.com)",
  "`inline equation: $E=mc^2$`",              "inline equation: $E=mc^2$",
  "`image: ![]()`",                           "image: ![]()",
  "`# Biggest Heading`",                      "# Biggest Heading",
  "`###### Smallest Heading`",                
  "<span style='font-size:16px;'>Smallest Heading</span>",
  "`> block quote`",                          
  "<span style='padding-left:1rem;border-left-color:#999;border-left-style:solid;border-left-width:1px;'>block quote</span>",
  "`* unordered list`<br>`* item 2`<br>`   + sub-item 1`<br>`   + sub-item 2`",
  "* unordered list\n* item 2\n   + sub-item 1\n   + sub-item 2",
  
  "`1. ordered list`<br>`2. item 2`<br>`   + sub-item 1`<br>`   + sub-item 2`",
  "1. ordered list\n2. item 2\n   + sub-item 1\n   + sub-item 2",
) |>
  gt(caption = "Markdown Reference", id = "md_ref") |>
  fmt_markdown(columns = everything()) |>
  cols_label(
    a = "Write this down as Markdown...",
    b = "...get this in HTML form"
  ) |>
  tab_options(table.width = pct(100)) |>
  opt_css(
    css = "
    #md_ref code {
      color: #333333;
    }"
  )
```

## Visual Markdown Editor

emoji, spell checking, commenting
