# R Lists and Factors {#r_lists_and_factors}

```{r setup, include=FALSE, echo=FALSE}
library(dspatterns)
library(tidyverse)
```

This chapter covers

- Working with lists: creating, modifying, and accessing data from them
- Writing our own functions that either create list output, use a list as input, or both

When working in **R** long enough, you'll encounter list objects again and again. You may bump into factors every now are again but try to side-step their use. However, these are very important and useful data structures and we ought to fully understand them. This chapter goes over lists, which we've seen before, in much more detail. We'll understand how to better create lists, how to access data from them, how to combine lists together, and how to go to and from lists and data frames. Factors will be introduced in the second part of this chapter. They shouldn't be something to be avoided as they are genuinely useful, especially in the context of plotting data with **ggplot**.

We'll use a dataset called `german_cities` (available in the **dspatterns** package) to work through examples that involve factors, eventually getting to nicer and nicer plots. The functions that we'll use throughout this chapter will be an even mix of base **R** functions and functions that are available in the *Tidyverse* packages **forcats** and **ggplot**.

## Lists in **R**

A list object in **R** is kind of like a container that can hold any type of **R** object, even other lists. The components of the list can have names, they can be unnamed (relying on numerical indices to distinguish between the components), or, we can have a mixture of named and unnamed components. They can be as simple as a vector, or, have numerous nested components. A good enough analogy for an **R** list is a computer's file system, where directories might hold elements (including more directories) and it's the files that are the addressable elements. A well-structured list is valuable as a complex data structure. Given that a function can only return a single object, and that a function may create complex data, a list adorned with different types of information can be returned and further used by other functions that require a little or a lot of that information.

### Making Named Lists and Accessing Their Elements

Just as we can create a tibble with **dplyr**'s `tibble()` function and a vector with `c()`, a list object can be produced by using the `list()` function. We saw an example of this earlier in *Chapter 7*. That was an example of a named list, where each element has a name (letters from `a` to `e`). Now, here's a question: why use a named list? The names can serve as easy-to-use, easy-to-understand identifiers for what to expect inside! For example, let's take a hypothetical list called `baseball_stats.` We might have names for elements called `"batting"`, `"pitching"`, and `"fielding"`. If we do, it's easy for the user handling that list to get batting statistics by simply using `baseball_stats$batting` in their code.

Let's make a named list here with three wildly different elements: a numeric vector, a tibble, and a list.

`r dspatterns::code_hints(
"**CODE //** Creating a three-element, named list object."
)`
```{r named-list-of-3-elements, paged.print=FALSE}
a_list <- 
  list(
 a = c(1:5, 8:3, 2),
 b = tibble(x = 1:3, y = rep(TRUE, 3)),
 c = list(A = 1:3, B = LETTERS[20:26])
  )

a_list
```

Seeing this printed out produces a correspondingly large amount of output.

Interestingly, the same rules that apply to tibbles also apply to list when it comes to accessing individual elements. We can use the `$` and `[[` operators to extract a specific element from the list object. The printed output provides clues on how to access the vectors of the list of element `c` (i.e., `$c$A` and `$c$B`). So, `a_list$c$A` gives us `[1] 1 2 3` in the console, `a_list$a[9:11]` results in `[1] 5 4 3`, and `a_list$b$y` yields `[1] TRUE TRUE TRUE`.

If working with a more complicated list, a nicer way of examining the structure of it is by using the `str()` function:

`r dspatterns::code_hints(
"**CODE //** Examining the structure of a list object with ~~str()~~."
)`
```{r structure-of-a-list}
str(a_list)
```

This output is easier to parse than simply printing the list object. In just a few lines we can see the hierarchy of the list and, importantly, get the names of the individual elements. Should you need just the names of the list components, we can use the `names()` function.

`r dspatterns::code_hints(
"**CODE //** Getting the names of the list components with ~~names()~~."
)`
```{r names-in-a-list}
names(a_list)
```

It was noted that the `[[` operator can be used to obtain a slice of the data. Let's quickly compare the ways in which we can get the same data extracts from different uses of operators:
- `a_list$c$A` or `a_list[["c"]][["A"]]` or `a_list[[3]][[1]]`
- `a_list$a[9:11]` or `a_list[["a"]][9:11]` or `a_list$a[[1]][9:11]`
- `a_list$b$y` or `a_list[["b"]][["y"]]` or `a_list[[2]][[2]]`

The three different methods for writing these subsetting expressions are (1) `$` with element names, (2) `[[` with element names in quotes, and (3) `[[` with element indices. You may ask why there are so many methods for getting at the same result. Well, the inputs are different. With the `$` notation the name is required, and it is essentially hardcoded (we can't use a variable like we might with the `[[` notation. The use of the `$` is great for interactive analysis (less typing!) or situations where we don't need a variable to get the name of the element. The use of `[[` works for both names and indices, and, we can use variables to access the elements of a list (imagine some code that produces the variable with the right index and then using that variable in `[[` to get the intended list element).

To help eliminate errors and reduce guesswork, the **RStudio** IDE provides autocomplete support for lists (and other objects) that works quite well with the `$` operator. Try creating the `a_list` object and typing `a_list$`, a contextual menu will appear with the available list elements at that juncture (use the up and down arrows to navigate and <ENTER> to select an element).

Accessing elements inside of lists is really something you need to experiment with and practice. It's suggested that you try out the above statements and change the input statements until you get a feel for the behavior. For example, here is something that constantly trips people up (myself included): what is the difference between `a_list[1]` and `a_list[[1]]`? Let's try both incantations in the next two code listings.

`r dspatterns::code_hints(
"**CODE //** Getting the first element of ~~a_list~~ with ~~[ ]~~."
)`
```{r a-list-element-single-brackets}
a_list[1]
```

`r dspatterns::code_hints(
"**CODE //** Getting the first element of ~~a_list~~ with ~~[[ ]]~~."
)`
```{r a-list-element-double-brackets}
a_list[[1]]
```

From both, we see similar output... what's the difference? The first statement, `a_list[1]`, gets the container for the vector whereas the second statement, `a_list[[1]]`, gets us the vector itself. We can sanity check this with `class(a_list[1])` and `class(a_list[[1]])`, which return `[1] "list"` (because it's the enclosing part of the vector) and `[1] "numeric"` (we are accessing the actual vector, which is numeric).

### Working with Unnamed Lists

Let's take a look at an unnamed list and see how things change without the use of names. Within the next code listing we'll create the `b_list` list object, which is similar to the `a_list` object introduced in the first code listing except that the top level of the list doesn't have the names `"a"`, `"b"`, or `"c"` (it has no names whatsoever).

`r dspatterns::code_hints(
"**CODE //** Creating a three-element, unnamed list object."
)`
```{r unnamed-list-of-3-elements, paged.print=FALSE}
b_list <- 
  list(
 c(1:5, 8:3, 2),
 tibble(x = 1:3, y = rep(TRUE, 3)),
 list(A = 1:3, B = LETTERS[20:26])
  )

b_list
```

The names are indeed gone, and, in their place, we get index values in double square brackets (`[[1]]`, `[[2]]`, and `[[3]]`). The output still indicates how to access elements of the unnamed list. To get the second (named!) character vector in the list of element `[[3]]`, we would use `b_list[[3]]$B`. To get the fourth element of that nested vector, it's `b_list[[3]]$B[4]`. Definitely take the time to practice accessing all of the elements of `a_list` and `b_list` until it becomes second nature. It's all worth doing because it helps build an intuition on how lists are structured and how to extract data.

We ought to use the `str()` function often to get a sense of the overall structure of a list (especially one that's foisted upon us, perhaps as output from a function). So, let's do that for the `b_list` object:

`r dspatterns::code_hints(
"**CODE //** Examining the structure of the ~~b_list~~ object with ~~str()~~."
)`
```{r structure-of-the-b-list}
str(b_list)
```

Looks almost identical to the output from using `str(a_list)` except that the names that would normally appear before the leftmost colons are not there! Certainly, it would be nicer to have numbers there but instead we need to count through these top-level elements ourselves (easy for small lists, increasingly more difficult for those larger ones). There really aren't any names and, indeed, using `names(b_list)` gives us `NULL`.

### Modifying Elements of Lists

You got this fancy list object and you took the time to get it just right. You should be proud because list-making can be a bit more involved than creating vectors or tibbles. The time will come, however, when you need to modify one or more elements of the list. No problem, we'll now explore how to change a part of a list while keeping the other components untouched.

Want to give that unnamed list some names? We can use the `names()` function to do that. Just ensure that the number of names provided is the same length as the list (check that with the `length()` function).

`r dspatterns::code_hints(
"**CODE //** Giving the unnamed list (~~b_list~~) some names."
)`
```{r b-list-gets-names, paged.print=FALSE}
names(b_list) <- c("a", "b", "c")

b_list
```

The output of `b_list` is now exactly the same as printing out the `a_list` object.

Perhaps we'd like to mutate the tibble inside the list, adding a character column named `z` with the elements `"a"`, `"b"`, and `"c"`. The key to making this work is having the right type of assignment: we need to assign into the list. This of course requires that we operate on the right subset of the list. For both, the answer is lies in using `b_list$b`.

`r dspatterns::code_hints(
"**CODE //** Modifying the tibble component of ~~b_list~~ requires accessing the ~~b_list$b~~ component."
)`
```{r modifying-the-b-list-tibble, paged.print=FALSE}
b_list$b <- 
  b_list$b %>%
  mutate(z = c("a", "b", "c"))

b_list
```

Notice now that the `b_list` object has its tibble in the second element updated with the change, keeping the first and third elements unchanged.

Setting values by element is possible for other data types as well (e.g., vectors, data frames, and tibbles). So long as you assign into the correct part of the data structure (preferably with the correct number of values), existing objects can be modified with great precision.

### Transforming Lists

The **R** list object is quite malleable in that it can be combined in numerous ways with other objects, transformed into entirely different data structures, and enable some pretty complex yet useful transformations. Let's walk through a few different list transformations of lists with examples, explanations, and possible use cases.

#### The `c()` Function
	
It's possible to combine lists with the `c()` function. This might be useful in the context of a function where you have a list as the main input (i.e., the data object to be manipulated), and you want to add elements to that list (before returning that enhanced list). 

We know that vectors can be combined together into a larger vector; lists are another data type that can undergo combining with `c()`. We can try creating the a_list object by combining three named lists, and we'll succeed if we use `c()`.

`r dspatterns::code_hints(
"**CODE //** Combining three named lists into a larger list with ~~c()~~ (reproducing the ~~a_list~~ object)."
)`
```{r combining-lists-with-c, paged.print=FALSE}
a_1 <- list(a = c(1:5, 8:3, 2))
a_2 <- list(b = tibble(x = 1:3, y = rep(TRUE, 3)))
a_3 <- list(c = list(A = 1:3, B = LETTERS[20:26]))

a_list <- c(a_1, a_2, a_3)

a_list
```

The output of this `a_list` is the same as the `a_list` object created in the first code listing.

If we don't want a certain component of the list, assign `NULL` to it. It's like sending that component into an **R** oblivion. Let's get rid of the tibble in `a_list` with this type of assignment:

`r dspatterns::code_hints(
"**CODE //** Removing the second element of the ~~a_list~~ object using ~~NULL~~."
)`
```{r removing-tibble-with-NULL, paged.print=FALSE}
a_list$b <- NULL

a_list
```

Yes, the second element has been removed. We could very well use numbers as well to perform the removal with `NULL.` Let's try that on `a_list` within the next code listing.

`r dspatterns::code_hints(
"**CODE //** Removing the first element of the ~~a_list~~ object using ~~NULL~~ (this time with an index value)."
)`
```{r removing-element-1-with-NULL, paged.print=FALSE}
a_list[1] <- NULL

a_list
```

We are now left with a list-in-a-list that was once the third element in `a_list` (it's now the first and only top-level element).

#### Transformations Between Data Frames and Lists

Data frames and tibbles are lists. The `data.frame()` documentation page (accessed by using `?data.frame`) states: "a data frame is a list of variables of the same number of rows with unique row names, given class 'data.frame'  (with a few restrictions in place)". We can convince ourselves of this by using the base **R** `typeof()` function with data frames and tibbles. Invoked with empty versions of each, using `typeof(data.frame())` and `typeof(tibble())`, we get `"list"` returned in both cases (just as with `typeof(list())`).

Let's create a list that can potentially become a data frame or tibble. The list will be named (these become the column names), each element of the list will be a single vector, and all vectors will have the same length.

`r dspatterns::code_hints(
"**CODE //** A named list that is suitable for transformation to a data frame or tibble."
)`
```{r list-to-transform-to-table, paged.print=FALSE}
list_df <-
  list(
    col_1 = 4:6,
    col_2 = c("x", "y", "z")
  )

list_df
```

This is a list that conforms to the rules, let's use the base **R** `as.data.frame()` function to make a data frame from it. There's a catch though! We need to use `stringsAsFactors = FALSE` (just as we do with `data.frame()`) so that the character vector, `col_2`, isn't coerced to a factor.

`r dspatterns::code_hints(
"**CODE //** Transforming the ~~list_df~~ list into a data frame with ~~as.data.frame()~~."
)`
```{r transform-as-data-frame, paged.print=FALSE}
df_from_list <- as.data.frame(list_df, stringsAsFactors = FALSE)

df_from_list
```

The way that the data frame is printed out makes it difficult to determine whether `col_2` is a character- or a factor-based column (try leaving out `stringsAsFactors = FALSE` in the above code, the resulting `df_from_list` object prints exactly the same). For this reason, among a few others, making a tibble with **dplyr**'s `as_tibble()` function is preferred since it would be obvious, when printed, what our column types are (plus, it doesn't create factor columns from character vectors in the first place). The next code listing is a variation on the last one, this time using **dplyr**'s `as_tibble()` function to transform the list into a tibble.

`r dspatterns::code_hints(
"**CODE //** Transforming the ~~list_df~~ list into a tibble with ~~as_tibble()~~."
)`
```{r transform-as-tibble, paged.print=FALSE}
tbl_from_list <- as_tibble(list_df)

tbl_from_list
```

Finally, we can transform a tibble or data frame to a list with the `as.list()` function. This is useful when the contents of a table might be better served when structured as a list. One scenario is when the table contents serve as an adequate starting point for a list, where additional elements are to be added (e.g., vectors of different sizes, etc.). In the next code listing we'll see an example use of `as.list()` with the `tbl_from_list` tibble (though this works equally well with the `df_from_list` object).

`r dspatterns::code_hints(
"**CODE //** Transforming the tibble into a list with ~~as.list()~~."
)`
```{r transform-tibble-to-list, paged.print=FALSE}
list_from_tbl <- as.list(tbl_from_list)

list_from_tbl
```

Experimenting with these examples of list transformations provides us with useful techniques, especially as we get to using lists in more complex ways within functions. In **R** functions, sometimes a list makes sense and, other times, a table (like a tibble or data frame) can be better. Moving between the two data structures is sometimes necessary and that's where these skills come into play.

### Creating Functions that Involve Lists

Lists are useful as both output and inputs for functions. While we can't return two objects at the same time from a function we can, however, do the next best thing: return a list that's structured with as much useful output as necessary. A list that serves as a primary input is a time-tested practice; we can use the list object as the main data object and transform it in interesting ways through functions written for transformation tasks.

Let's make a function that takes a numeric vector returns a list object. **R** is great at doing statistical calculations so this new function will help us obtain descriptive statistics (mean, min, max, etc.) for a numeric vector, putting those values into a list. We'll create the function named `get_descriptive_stats()` in the following code listing and then immediately test it out using a numeric vector created with `c(2.3, 8.1, 5.5, 3.9, 6.0, 2.1, 8.5)`.

`r dspatterns::code_hints(
"**CODE //** Example of a function (~~get_descriptive_stats()~~) that returns a list object."
)`
```{r get-descriptive-stats}
get_descriptive_stats <- function(x, na.rm = TRUE) {
  
  list(
    values = x,
    mean = mean(x, na.rm = na.rm),
    sd   = sd(x, na.rm = na.rm),
    min  = min(x, na.rm = na.rm),
    max  = max(x, na.rm = na.rm),
    rank = rank(x)
  )
}

stats <- get_descriptive_stats(c(2.3, 8.1, 5.5, 3.9, 6.0, 2.1, 8.5))

stats
```

Again, what's nice about the list returned from `get_descriptive_stats()` is that you get a useful bundle of information in the stats object. If you should need the mean of the input values, you can then use `stats$mean`. The range, on the other hand, can be calculated with `stats$max – stats$min`.

We can design functions that use a list object as input. As a very simple example, we can take the output from `get_descriptive_stats()` and augment it with additional data (let's use the 25th, 50th, and 75th percentiles). In the next code listing we define a new function called `add_percentiles()`, which expects a list that is returned by `get_descriptive_stats()`. As before, we'll use the function right away in the example and observe that the list output (`stats_extra`) has the extra components we wanted from the function.
	
`r dspatterns::code_hints(
"**CODE //** Example of a function (~~add_percentiles()~~) that takes a list object as input and returns an augmented version of the input list."
)`
```{r add-percentiles}
add_percentiles <- function(stats_list, na.rm = TRUE) {
  
  x <- stats_list$values
  
  c(
    stats_list,
    list(
      p25 = unname(stats::quantile(x, probs = 0.25, na.rm = na.rm)),
      p50 = unname(stats::quantile(x, probs = 0.50, na.rm = na.rm)),
      p75 = unname(stats::quantile(x, probs = 0.75, na.rm = na.rm))
    )
  )
}

stats_extra <- add_percentiles(stats_list = stats)

stats_extra
```

The output looks good! Did you notice the use of the `c()` function within the body of `add_percentiles()`? It's definitely a great way to combine two lists together.

## Summary

- Creating a list object is done with the `list()` function; lists can be named, unnamed, or be partially named
- We can use the `$` and `[[` operators to access list elements or to do assignment on elements of a list
- The `str()` function can be used with a list object to concisely show us its structure; the `names()` function is also useful for list inspection: it gives us its names (or lack thereof)
- We can combine lists with the `c()` function, remove list elements by carefully assigning `NULL` to those elements, transform lists to data frames or tibbles with `as.data.frame()` or `as_tibble()`, or, do the opposite with `as.list()`
- Creating functions that return list objects is useful in circumstances where complex information is needed with a single function call (because we can't return multiple objects, but a list can hold disparate object types)
- Using lists as an input for a function can be a good idea should we need a complex input or if we want to further modify the one or more components of that list
- Factors are vectors that contain levels: integers that describe the ordering of the factor's values; we can inspect factors with a number of useful base **R** functions (`as.integer()`, `levels()`, `nlevels()`, etc.) and the `fct_count()` function from the **forcats** package
- Should factors vectors get modified by changing the number of elements, the storage of factor levels is not changed (and we may have levels for nonexistent values): remedy this with `fct_drop()`
- We can get fine control over plotting categorical data when using factor transformation functions from **forcats** just before using **ggplot**; these **forcats** functions were useful in this regard: `fct_reorder()`, `fct_rev()`, `fct_recode()`, `fct_infreq()`, `fct_inseq()`, and `fct_lump_n()`.
